HO unification from object language to meta language

### 2024-02-23 paper/main.tex
labels

\begin{coqcode}
  Check _ : Decision (forall y: fin 7, nfact y 3).  (* g *)
\end{coqcode}

@@

\begin{elpicode}
finite (app[c"fin", N]).                             % r1
decision (app [c"nfact", N, NF]).                    % r2
decision (all A x\ app[P, x]) :- finite A,           % r3
  pi w\ decision (app[P, w]).
\end{elpicode}

\noindent
Unfortunately this direct translation of rule (r3) uses the
predicate \coqIn{P} essentially as a first order term for the meta
language (its type is \elpiIn{tm}).
If we try to backchain the rule (r3) on the encoding of the goal (g) below:

\begin{elpicode}
decision (all (app[c"fin", c"7"]) y\                 % g
  app[c"nfact", y, c"3"]).
\end{elpicode}

\noindent
we fail because this ``higher order'' unification problem (for DTT)
is phrased as a first-order unification problem in the meta language:
the two lists of terms have different lengths!

\begin{elpicode}
app[c"nfact", y, c"3"] = app[P, y]                   % p
\end{elpicode}

\noindent
In this paper we study a more sophisticated encoding of Coq terms allowing
us to rephrase the problematic rule (r3) as follows:

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm A P, finite A,   % r3a
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
Since \elpiIn{Pm} is an higher-order unification variable with \elpiIn{x}
in its scope, the unification problem (pa) admits one solution:

\begin{elpicode}
app[c"nfact", y, c"3"] = Pm y                       % pa
Pm = x\ app[c"nfact", x, c"3"]       % assignment for Pm
A = app[c"fin", c"7"]                % assignment for A
\end{elpicode}

\noindent
After unifying the head of rule (r3a) with the goal Elpi runs
the premise \elpiIn{link Pm A P} that is in charge of bringing the
assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
back to the domain of Coq terms (the type \elpiIn{tm}):

@@

\noindent
In turn this redex prevents the rule (r2) to backchain properly since
the following unification problem has no solution:

@@

\noindent
One could ignore this similarity, and ``just'' describe the object language
unification procedure in the meta language, that is crafting a \elpiIn{unif}
predicate to be used as follows in rule (r3):

@@

Our choice of encoding of DTT may look weird to the reader familiar with
LF, since used a shallow encoding of classes and binders, but not of the
``lambda calculus'' part of DTT. Here a more lightweight encoding
that unfortunately does not fit our use case

\begin{elpicode}
finite (fin N).
decision (nfact N NF).
decision (all A x\ P x) :- 
  (pi x\ decision (P x)), finite A.
\end{elpicode}

but in DTT this is not always possible and not handy in our use case,
since the arity of constants is not fixed.

\begin{coqcode}
Fixpoint narr T n := 
  if n is S m then T -> narr T m else T.
Definition nsum n : narr nat (n+1).
Check nsum 2   8 9 : nat.
Check nsum 3 7 8 9 : nat.
\end{coqcode}

moreover we use the same encoding for meta programming, or even just to provide
hand written rules. We want to access the syntax of OL, so our embedding cannot
be that shallow. We want to keep it shallow for the binders, but we need
the c, app and lam nodes. % all ?

Note that this~\cite{felty93lics} is related and make the
discrepancy between the types of ML and DTT visible. In this case
one needs 4 application nodes. Moreover the objective is an encoding
of terms, proofs, not proof search. Also note the conv predicate,
akin to the unif we rule out.

This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.

>>>>

\begin{coqcode}
  Check _ : Decision (forall y: fin 7, nfact y 3).   ~\customlabel{goal:g}{(g)}~
\end{coqcode}

@@

\begin{elpicode}
finite (app[c"fin", N]).                              ~\customlabel{clause:r1}{(r1)}~
decision (app [c"nfact", N, NF]).                     ~\customlabel{clause:r2}{(r2)}~
decision (all A x\ app[P, x]) :- finite A,            ~\customlabel{clause:r3}{(r3)}~
  pi w\ decision (app[P, w]).
\end{elpicode}

\noindent
Unfortunately this direct translation of rule \ref{clause:r3} uses the
predicate \coqIn{P} essentially as a first order term for the meta
language (its type is \elpiIn{tm}).
If we try to backchain the rule \ref{clause:r3} on the encoding of the goal \ref{goal:g} below:

\begin{elpicode}
decision (all (app[c"fin", c"7"]) y\                 
  app[c"nfact", y, c"3"]).
\end{elpicode}

\noindent
we obtain an unsolvable unification problem \ref{problem:p}:
the two lists of terms have different lengths! The root cause is that
\ref{problem:p} is an higher order in DTT, but becomes
first order due to the encoding in the meta language.

\begin{elpicode}
app[c"nfact", y, c"3"] = app[P, y]                     ~\customlabel{problem:p}{(p)}~
\end{elpicode}

\noindent
In this paper we study a more sophisticated encoding of Coq terms allowing
us to rephrase the problematic rule \ref{clause:r3} as follows:

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm A P, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
Since \elpiIn{Pm} is an higher order unification variable with \elpiIn{x}
in its scope, the unification problem \ref{problem:pa} admits one solution:

\begin{elpicode}
app[c"nfact", y, c"3"] = Pm y                         ~\customlabel{problem:pa}{(pa)}~
Pm = x\ app[c"nfact", x, c"3"]       % assignment for Pm
A = app[c"fin", c"7"]                % assignment for A
\end{elpicode}

\noindent
After unifying the head of rule \ref{clause:r3a} with the goal Elpi runs
the premise \elpiIn{link Pm A P} that is in charge of bringing the
assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
back to the domain of Coq terms (the type \elpiIn{tm}):

@@

\noindent
In turn this redex prevents the rule \ref{clause:r2} to backchain properly since
the following unification problem has no solution:

@@

\noindent
One could ignore this similarity, and ``just'' describe the object language
unification procedure in the meta language, that is crafting a \elpiIn{unif}
predicate to be used as follows in rule \ref{clause:r3}:

@@

Our encoding of DTT may look ``semi shallow'' since we we use the meta language
lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
A fully shallow encoding unfortunately does not fit our use case, although
it would make the running example work:

\begin{elpicode}
finite (fin N).
decision (nfact N NF).
decision (all A x\ P x) :- finite A, pi x\ decision (P x).
\end{elpicode}

\noindent
There are two reasons for dismissing this encoding. The first one is that
in DTT it is not always possible to adopt it since the type system
of the meta language is too weak to accommodate terms with a variable arity,
like the following example:

\begin{coqcode}
Fixpoint arr T n := if n is S m then T -> arr T m else T.
Definition sum n : arr nat n := ...
Check sum 2   7 8   : nat.
Check sum 3   7 8 9 : nat.
\end{coqcode}

\noindent
The second reason is the encoding for Coq is used for meta programming the
system, hence it must accommodate the manipulation of terms that are now
know in advance (not even defined in Coq) without using introspection
primitives such as Prologs's \texttt{functor} and \texttt{arg}.

In the literature we could find a few related encoding of DTT.
TODO In~\cite{felty93lics} is related and make the
discrepancy between the types of ML and DTT visible. In this case
one needs 4 application nodes. Moreover the objective is an encoding
of terms, proofs, not proof search. Also note the conv predicate,
akin to the unif we rule out.

TODO This other paper~\cite{10.1007/978-3-031-38499-8_25} should also be cited.

None of the encodings above provide a solution to our problem.
